The Evolution of Phenotypically Invariant Genetic Networks

### 2017-09-11 kalman_walk_paper.tex
edited examples

The molecular machinery, interacting with the environment, and bridging genotype to phenotype
can be mathematically described as a dynamical system -- or a system of differential equations \citep{jaeger2015comet}.
 Movement in this direction is ongoing, as researchers have begun to study 
the evolution of both abstract \citep{wagner1994evolution, wagner1996does,  siegal2002waddington, bergman2003evolutionary, draghi2015robustness} and empirically inspired computational and mathematical models of gene regulatory networks (GRNs) \citep{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative, chertkova2017insilico}. If we allow the reasonable assumption that the genotype-phenotype map can be represented as a system of differential equations, we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.

@@

Organisms' phenotypes are constructred by gene by gene by environment interactions. Here we simply define the \emph{phenotype} to be the temporal molecular dynamics directly under natural selection. The \emph{what}, \emph{when}, and \emph{how much}, of an organism's molecules that are physiologically or otherwise relevant to survival.

@@

    Suppose gene-2 up-regulates the transcription of gene-1 and that gene-1 down-regulates gene-2 with equal magnitudes, whose concentrations are given by $\kappa_1(t)$ and $\kappa_2(t)$. Furthermore, suppose that only the dynamics of gene-1 are consequential to the cell cycle (perhaps the amount of gene-1 activates another downstream gene network). Lastly suppose that the production of both genes is stimulated by an impulse of a molecule present immediately after division.

    If the rate each of these genes is expressed is a linear function of their concentrations, the dynamics of the system are given by
    \begin{align*}
      \dot{\kappa_{1}}(t) &= \kappa_{2}(t) + u(t) \\
      \dot{\kappa_{2}}(t) &= - \kappa_{1}(t) + u(t)
    \end{align*}
    where $\dot{\kappa}$ denotes the time derivative. The initial conditions $\kappa_{1}(0)$ and $\kappa_{2}(0)$, and the input $u(t)$ then determine the concentrations through time. If we record the regulatory coefficients in the matrix
    \begin{align*}
      A &= \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} ,
    \end{align*}
    and define the column vector $B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}$, then in matrix notation the dynamics are
      \begin{align*}
      \dot{\kappa} &= A \kappa(t) + B u(t) .
      \end{align*}
      Since only the dynamics of gene-1 are directly relevant to biological function, the dynamics of interest are given by
      \begin{align*}
      \phi(t) &= C \kappa (t)
      \end{align*}
      where the row vector $C$ is defined as $C = \begin{bmatrix} 1 & 0 \end{bmatrix}$. 
  %  \begin{align*}
  %    A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
  %  \end{align*}
 %       The oscillatory system $\Sigma$ is thus given as
 %   \begin{align*}
 %     \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
 %       0 & 1 \\ 
 %      -1 & 0 
 %       \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
 %         \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
 %    \end{align*}

      Since the input is simply an impulse, its phenotype is equivalent to its impulse response
      \begin{align*}
        \phi(t) = h(t) = \sin(t) + \cos(t)  .
      \end{align*}

    \begin{figure}[H]
       % \begin{center} 
      \centering
         \begin{tabular}{cc}
            \begin{tikzpicture}
            \begin{scope}[every node/.style={circle,thick,draw}]
                \node (A) at (0,0) {$\kappa_{1}$};
                \node[dashed] (B) at (3,0) {$\kappa_{2}$};
                \node[shape=rectangle] (U) at (1.5,2) {input};
                \node[shape=rectangle] (y) at (1.5,-2) {output};
            \end{scope}

@@

            \end{tikzpicture} &
       % \end{center}
      %  \caption{
      %      Diagram of Example \ref{ex:osc} in the text.
      %      \plr{explain what arrows mean if nec}
      %      \label{fig:oscillator_diagram}}
   % \end{figure}
   % \begin{figure}[H]
    %  \centering
      \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_impulse}
   \end{tabular}
  %  \end{center}
      \caption{(Left) Graphical representation of the cell cycle control gene network, and (right) plot of the phenotype $\phi(t)$ against time $t$.}
    \end{figure}
      We return to the evolution of such a system below.
  \end{example}

@@

%\plr{This is not part of the definition of $\mathcal{I}$:}
%\js{should we delete this example ($A(0) \times A(2)$ hybrids)? I like how clean it is -- but I'm not sure it's that important in this context. It's a good proof of principle though.}
%\jss{$F_1$s created by crossing phenotypically equivalent oscillators $A(0)$ and $A(2)$ have a phenotype of $\phi_{F_{1}}(t) = e^{t}$, 
%in contrast to both parents, who have $\phi_{P_{1}}(t) = \phi_{P_{2}}(t) = \sin(t) + \cos(t)$. 
%The hybrid phenotype is significantly different (it does not oscillate and increases infinitely) despite the phenotypic equivalence of the parents.
%%\plr{don't need these to be big equations.}
%%\begin{align*}
%%  \fit \left(\phi_{P_{1}}\right) = \fit \left( \phi_{P_{2}} \right) &= 1 \\
%%  \fit\left(\phi_{F_{1}}\right) &= 0 \\
%%  \mathcal{I} &= 0
%%\end{align*}
%
%%Thus if populations $1$ and $2$ are homogeneous $A(0)$ and $A(2)$, respectively, we say that they are completely incompatible as $\mathcal{I} = 0$.
%
%  \begin{figure}[H]
%    \centering
%    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{expF1}
%    \caption{$F_1$ hybrid (orange) and parental (blue) phenotypic oscillator dynamics for an $A(0)$ by $A(2)$ cross. The hybrid fails to oscillate and exhibits qualitatively different dynamics.}
%  \end{figure}
%  }
%%     \begin{example}[F1 Reproductive Incompatibility in an Oscillating Gene Network]
%%       DMI examples...
%%       \begin{figure}[H]
%%         \includegraphics[width=0.5\textwidth]{A2_A3_osci_F1_hyb}
%%       \end{figure}
%%     \end{example}
%
%%     \begin{example}[Not all Networks can Host Incompatibilities]
%%       convex sets cant have DMIs
%
%%       \begin{align*}
%%         h(t) = 2 e^{- \theta t}
%%       \end{align*}
%%       Any non-minimal system with rows summing to $\theta$ is PI. Further, these systems are closed under averaging (mating) and row swapping (meiosis), leaving all hybrids optimally fit. The set of gene matrices is affine and therefore convex.  
%%     \end{example}
%
%
%\plri{Missing: a statement about when there are neutral directions, and how many.}
%
%\plr{I think we should first put the section about quantitative drift in, and then the following section that tries to get at real values of parameters.}
%
%
 \begin{example}[All Phenotypically Equivalent Cell Cycle Control Networks] \label{ex:all_osc}

%\plr{this is same as above, so reference instead of writing out again?}
%    Let
%    \begin{align*}
%      A(0) = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix}, \qquad B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix},
%    \end{align*}
%    with $V(\tau)$ preserving both $B$ and $C$, then
    The set of all two-gene regulatory networks phenotypically equivalent to the cell cycle control network in \ref{ex:oscillator}, where only $A$ can vary, is given by
    \begin{align*}
      A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix}  \qquad \forall \ \tau \neq 1 %\\
     %   B &= \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix} \\
       %   h(t) &= \sin(t) + \cos(t) \qquad \forall \ \tau \neq 1
    \end{align*}
    \begin{figure}[H]
    \centering
% Nice tikz!!!
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$\kappa_{1}$};
    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
    \node[shape=rectangle] (U) at (2,2) {input};
    \node[shape=rectangle] (y) at (2,-2) {output};
\end{scope}

@@

 % \begin{example}[External Equivalence does not imply internal equivalence]
   Despite the phenotypic equivalence of all instantiations of $A(\tau)$, the kryptotypes, vary as a function of $\tau$. 
    Gene-1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the dynamics of gene-2 (orange) differ with $\tau$.
  \begin{figure}[H]
    \centering
    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_A0_A2_both_compare}
    \caption{Gene-1 (blue) and gene-2 (orange) dynamics for $A(0)$ (top) and $A(2)$ (bottom). Both (top and bottom) gene-1 dynamics are given by $ \kappa_{1} = \sin(t) + \cos(t)$, and gene-2 by $\kappa_{2} = \cos(t) - \sin(t)$ (top) and $\kappa'_{2} = \cos(t) + 3 \sin(t)$ (bottom).}
  \end{figure}
\end{example}

>>>>

The molecular machinery, interacting with the environment, and bridging genotype to phenotype
can be mathematically described as a dynamical system -- or a system of differential equations \citep{jaeger2015comet}.
 Movement in this direction is ongoing, as researchers have begun to study 
the evolution of both abstract \citep{wagner1994evolution, wagner1996does,  siegal2002waddington, bergman2003evolutionary, draghi2015robustness} and empirically inspired computational and mathematical models of gene regulatory networks \citep{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative, chertkova2017insilico}. If we allow the reasonable assumption that the genotype-phenotype map can be represented as a system of differential equations, we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.

@@

Organisms' phenotypes are constructred by gene by gene and gene by environment interactions. Here we simply define the \emph{phenotype} to be the temporal molecular dynamics directly under natural selection. The \emph{what}, \emph{when}, and \emph{how much}, of an organism's molecules that are physiologically or otherwise relevant to survival.

@@

    Suppose gene-2 up-regulates the transcription of gene-1 and that gene-1 down-regulates gene-2 with equal magnitudes, whose concentrations are given by $\kappa_1(t)$ and $\kappa_2(t)$. Furthermore, suppose that only the dynamics of gene-1 are consequential to the cell cycle (perhaps the amount of gene-1 activates another downstream gene network). Lastly suppose that the production of both genes is equally stimulated by an impulse of a molecule present immediately after division.

    If the rate each of these genes is expressed is a linear function of their concentrations, the dynamics of the system is given by
    \begin{alignat*}{2}
      %\left\{ \begin{array}{ll} 
      \dot{\kappa_{1}}(t) &= &&\kappa_{2}(t) + u(t) \\
        \dot{\kappa_{2}}(t) &= - &&\kappa_{1}(t) + u(t) \\
        \phi(t) &= &&\kappa_{1}(t)   .
      %\end{array} \right.
    \end{alignat*}
Equivalently, in matrix form the system regulatory coefficients are given as,
    $A \!=\! \left[\begin{smallmatrix} 0 & 1 \\ -1 & 0 \end{smallmatrix}\right] ,\ B \! =\! \left[\begin{smallmatrix} 1 \\ 1 \end{smallmatrix}\right] , \ C \! =\! \left[\begin{smallmatrix} 1 & 0 \end{smallmatrix}\right]$
  %  \begin{align*}
  %    A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
  %  \end{align*}
 %       The oscillatory system $\Sigma$ is thus given as
 %   \begin{align*}
 %     \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
 %       0 & 1 \\ 
 %      -1 & 0 
 %       \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
 %         \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
 %    \end{align*}

      Since the input is simply an impulse, its phenotype is equivalent to its impulse response,
      \begin{align*}
        \phi(t) = h(t) = \sin(t) + \cos(t)  .
      \end{align*}

    \begin{figure}[H]
       % \begin{center} 
      \centering
         \begin{tabular}{cc}
            \begin{tikzpicture}
            \begin{scope}[every node/.style={circle,thick,draw}]
                \node (A) at (0,0) {$\kappa_{1}$};
                \node[dashed] (B) at (3,0) {$\kappa_{2}$};
                \node[shape=rectangle] (U) at (1.5,1.5) {input};
                \node[shape=rectangle] (y) at (1.5,-1.5) {output};
            \end{scope}

@@

            \end{tikzpicture} &
      \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{osc_impulse}
   \end{tabular}
  %  \end{center}
      \caption{(Left) Graphical representation of the cell cycle control gene network, and (right) plot of the phenotype $\phi(t)$ against time $t$.}
    \end{figure}
      We return to the evolution of such a system below.
  \end{example}

@@

%\plr{This is not part of the definition of $\mathcal{I}$:}
%\js{should we delete this example ($A(0) \times A(2)$ hybrids)? I like how clean it is -- but I'm not sure it's that important in this context. It's a good proof of principle though.}
%\jss{$F_1$s created by crossing phenotypically equivalent oscillators $A(0)$ and $A(2)$ have a phenotype of $\phi_{F_{1}}(t) = e^{t}$, 
%in contrast to both parents, who have $\phi_{P_{1}}(t) = \phi_{P_{2}}(t) = \sin(t) + \cos(t)$. 
%The hybrid phenotype is significantly different (it does not oscillate and increases infinitely) despite the phenotypic equivalence of the parents.
%%\plr{don't need these to be big equations.}
%%\begin{align*}
%%  \fit \left(\phi_{P_{1}}\right) = \fit \left( \phi_{P_{2}} \right) &= 1 \\
%%  \fit\left(\phi_{F_{1}}\right) &= 0 \\
%%  \mathcal{I} &= 0
%%\end{align*}
%
%%Thus if populations $1$ and $2$ are homogeneous $A(0)$ and $A(2)$, respectively, we say that they are completely incompatible as $\mathcal{I} = 0$.
%
%  \begin{figure}[H]
%    \centering
%    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{expF1}
%    \caption{$F_1$ hybrid (orange) and parental (blue) phenotypic oscillator dynamics for an $A(0)$ by $A(2)$ cross. The hybrid fails to oscillate and exhibits qualitatively different dynamics.}
%  \end{figure}
%  }
%%     \begin{example}[F1 Reproductive Incompatibility in an Oscillating Gene Network]
%%       DMI examples...
%%       \begin{figure}[H]
%%         \includegraphics[width=0.5\textwidth]{A2_A3_osci_F1_hyb}
%%       \end{figure}
%%     \end{example}
%
%%     \begin{example}[Not all Networks can Host Incompatibilities]
%%       convex sets cant have DMIs
%
%%       \begin{align*}
%%         h(t) = 2 e^{- \theta t}
%%       \end{align*}
%%       Any non-minimal system with rows summing to $\theta$ is PI. Further, these systems are closed under averaging (mating) and row swapping (meiosis), leaving all hybrids optimally fit. The set of gene matrices is affine and therefore convex.  
%%     \end{example}
%
%
%\plri{Missing: a statement about when there are neutral directions, and how many.}
%
%\plr{I think we should first put the section about quantitative drift in, and then the following section that tries to get at real values of parameters.}
%
%
 \begin{example}[All Phenotypically Equivalent Cell Cycle Control Networks] \label{ex:all_osc}
    The set of all two-gene regulatory networks phenotypically equivalent to the cell cycle control network in \ref{ex:oscillator}, where only $A$ can vary, is given by
    %\begin{align*}
    %  A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix}  \qquad \forall \ \tau \neq 1 %\\
    %\end{align*}
    \begin{align*}
      A(\tau) &= \frac{1}{\tau-1} \begin{bmatrix} \tau & -1 \\ 2 \tau(\tau - 1) + 1 &  -\tau \end{bmatrix} \ \forall \ \tau \neq 1
    \end{align*}
    \begin{figure}[H]
    \centering
% Nice tikz!!!
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$\kappa_{1}$};
    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
    \node[shape=rectangle] (U) at (2,2) {input};
    \node[shape=rectangle] (y) at (2,-2) {output};
\end{scope}

@@

 % \begin{example}[External Equivalence does not imply internal equivalence]
   Despite the phenotypic equivalence of all instantiations of $A(\tau)$, the kryptotypes, vary as a function of $\tau$. 
    Gene-1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the dynamics of gene-2 (orange) differ with $\tau$.
  \begin{figure}[H]
    \centering
   % \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_A0_A2_both_compare}
    %\caption{Gene-1 (blue) and gene-2 (orange) dynamics for $A(0)$ (top) and $A(2)$ (bottom). Both (top and bottom) gene-1 dynamics are given by $ \kappa_{1} = \sin(t) + \cos(t)$, and gene-2 by $\kappa_{2} = \cos(t) - \sin(t)$ (top) and $\kappa'_{2} = \cos(t) + 3 \sin(t)$ (bottom).}
    \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{figures/osc_kryp_compare}
    \caption{(Gene-1 dynamics (blue) for both systems $A(0)$ and $A(2)$ are identical, however, $A(0)$ gene-2 dynamics (orange) differ from $A(2)$ (green). Both gene-1 dynamics are given by $\kappa_{1} = \sin(t) + \cos(t)$, and gene-2 by $\kappa_{2} = \cos(t) - \sin(t)$ ($A(0)$) and $\kappa'_{2} = \cos(t) + 3 \sin(t)$ ($A(2)$).}
  \end{figure}
\end{example}
